---
title: "Nouns and verbs"
author: "Martin Morgan, Herve Pages"
date: "June 19 - 20, 2017"
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
output: 
  BiocStyle::html_document
---

# _R_ data and programs

## Building blocks

```{r}
x <- rnorm(1000)
y <- x + rnorm(1000)
df <- data.frame(Independent = x, Dependent = y)
fit <- lm(Dependent ~ Independent, df)
anova(fit)
```

Atomic vectors

- `logical()`, `integer()`, `numeric()`, `complex()`, `raw()`, `character()`
- `list()`

Attributes

- Attributes in C

An implicit class -- `matrix()`

`list()` and `environment()`

- recursive
- `environment()`: reference semantics

Plain-old-functions

- First class objects

## S3

Class

- Definition -- list (usually) or atomic vector with `class` attribute
- Inheritance: _is-a_ and _has-a_
    - linear hierarchy

```{r}
People <- function(name = character(), age = character()) {
    stopifnot(
        is.character(name), is.numeric(age), length(name) == length(age)
    )
    structure(list(name=name, age=age), class = "People")
}

Employees <- function(people = People(), job = character()) {
    stopifnot(
        identical(class(people), "People"), is.character(job),
        length(people) == length(job)
    )
    employees <- people
    employees$job <- job
    class(employees) <- c("Employees", class(employees))
    employees
}

Company <- function(company_name = character(), empolyees = Employees()) {
    stopifnot(
        is.character(company_name), length(company_name) == 1L,
        !is.na(company_name),
        inherits(employees, "Employees")
    )
    structure(
        list(
            company_name = company_name, 
            employees = employees
        ),
        class = "Company"
    )
}
```

Generic

- 'Group' generics

Method

- 'Next' methods

Formal characteristics

- Strengths & weakneses

## S4

### Preamble

- Formal class definitions
- Many similar concepts, but greater flexibility in class
  _inheritance_ and method _dispatch_
- Despite it's reputation, not overly verbose compared to
  well-structured S3 classes

```{r}
People <- setClass(
    "People",
    slots = c(name = "character", age = "numeric")
)

setValidity("People", function(object) {
    if (!identical(length(name(object)), length(age(object)))) {
        "lengths of name() and age() differ"
    } else TRUE
})

Employees <- setClass(
    "Employees", 
    contains = "People",
    slots = c(job = "character")
)

setValidity("Employees", function(object) {
    if (!identical(length(employees(object)), length(jobs(object)))) {
        "length of employee() and job() differ"
    } else TRUE
})

Company <- setClass(
    "Company",
    slots = c(company_name = "character", employee = "Employees")
)

setValidity("Company", function(object) {
    if (length(company_name) != 1 || !is.na(company_name)) {
        "company_name must be length 1 and not NA"
    } else TRUE
})
```

## Classes, generics, and methods

Class

- Definition
- Inheritance
- Virtual classes
- Class unions
- Discovery, e.g., `getClass()`

Generic

- Argument versus signature

Method

- Dispatch

Formal characteristics

- Strengths & weakneses
- Column- versus row-oriented data

# Essential _Bioconductor_ classes

## Motivation / necessity

Users

Developers

Key classes for end users

- GenomicRanges
- SummariedExperiment

## Design in depth

Concepts

Efficiencies

Common 'tricks'
