---
title: "S3 classes and methods"
author: "Martin Morgan, Hervé Pagès"
date: "June 19 - 20, 2017"
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
output: 
  BiocStyle::html_document2
---



# _R_ data (classes) and programs (functions and methods)

## Building blocks

```{r}
x <- rnorm(1000)
y <- x + rnorm(1000)
df <- data.frame(Independent = x, Dependent = y)
fit <- lm(Dependent ~ Independent, df)
anova(fit)
```

Atomic vectors

- `logical()`, `integer()`, `numeric()`, `complex()`, `raw()`, `character()`

Attributes

- Attributes in C

An implicit class -- `matrix()`

`list()` and `environment()`

- recursive
- `environment()`: reference semantics

Plain-old-functions

- First class objects

## S3

Class

- Definition -- list (usually) or atomic vector with `class` attribute
- Inheritance: _is-a_ and _has-a_
    - linear hierarchy

```{r}
People <- function(name = character(), age = character()) {
    stopifnot(
        is.character(name), is.numeric(age), length(name) == length(age)
    )
    structure(list(name=name, age=age), class = "People")
}

Employees <- function(people = People(), job = character()) {
    stopifnot(
        identical(class(people), "People"), is.character(job),
        length(people) == length(job)
    )
    employees <- people
    employees$job <- job
    class(employees) <- c("Employees", class(employees))
    employees
}

Company <- function(company_name = character(), empolyees = Employees()) {
    stopifnot(
        is.character(company_name), length(company_name) == 1L,
        !is.na(company_name),
        inherits(employees, "Employees")
    )
    structure(
        list(
            company_name = company_name, 
            employees = employees
        ),
        class = "Company"
    )
}
```

Generic

- 'Group' generics

Method

- 'Next' methods

Formal characteristics

- Strengths & weakneses

### Exercises

**Exercise:** `length()` S3 method

`?length` indicates that this function is a (S3) generic. Define an S3
`length` method for the `People` class; note that we use this when
validating the input to `Employees()`.

**Exercise:** `name()` and `age()` accessors

The `Person` class has fields `name` and `age`. These are accessible
by reaching in to the implementation details and extracting the list
element, e.g., `Person()$name`. But this type of direct manipulation
is considered bad practice; it is better to have 'accessors' that
define the 'application programming interface (API)' for accessing
information from the object. Write plain-old-functions (e.g., like
`name <- function(x) ...`) to extract name and age.

**Exercise:** S3 generics and methods

Revise `name()` and `age()` to be S3 generics and methods. Reflect on
the beneifts and costs of plain-old-functions versus generic +
methods.

